---
title: "Students"
sidebar:
  style: "floating"
  contents:
    - Faculty.qmd
    - Students.qmd
    - Admissions.qmd
---

*"Boston University is an international, comprehensive, private research university, committed to educating students to be reflective, resourceful individuals ready to live, adapt, and lead in an interconnected world. Boston University is committed to generating new knowledge to benefit society."* BU Mission Statement

BU is an institution centered around students and education. Every year, a new cohort with new ideas and experiences matriculates, and every year a cohort of fresh content experts graduates. The constantly evolving student body guides the trajectory of the institution.

# Tuition and Aid

I was an undergraduate in 2022 when the "largest tuition hike in 14 years" was met with upset, frustration, and despair. However, this graph reveals what is essentially a linear increase in tuition. Of course, a 4% increase of an increasing significant amount of money hits students harder each year, but this graph shows maintenance of the status quo.

```{r stufin, message=FALSE, warning=FALSE}

suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(stringr)))
suppressMessages(suppressWarnings(library(tidyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggthemes)))
suppressMessages(suppressWarnings(library(hrbrthemes)))
suppressMessages(suppressWarnings(library(plotly)))
suppressMessages(suppressWarnings(library(scales)))

stu_aid <- read_csv("Stu.Aid.02.csv", show_col_types = FALSE)

stu_aid <- stu_aid %>%
  mutate(
    Year = as.integer(Year),
    `Average Aid` = as.numeric(`Average Aid`),
    `Percent Given Aid` = as.numeric(`Percent Given Aid`),
    `Total Tuition` = as.numeric(Total),
    Percent_Scaled = `Percent Given Aid` * 900
  )

long_data <- stu_aid %>%
  pivot_longer(cols = c("Average Aid", "Total Tuition", "Percent_Scaled"),
               names_to = "Metric",
               values_to = "Value") %>%
  mutate(Metric = case_when(
    Metric == "Average Aid" ~ "Average Aid",
    Metric == "Total Tuition" ~ "Total Tuition",
    Metric == "Percent_Scaled" ~ "Percent Given Aid"
  ))

p <- ggplot(long_data, aes(x = Year, y = Value, color = Metric, group = Metric,
                           text = paste("Year:", Year,
                                        "<br>Value:", 
                                        ifelse(Metric == "Percent Given Aid", 
                                               scales::percent(Value / 900, scale = 1), 
                                               scales::label_dollar()(Value))))) +
  geom_line(size = 1.2) +
  geom_point() +
  labs(title = "Student Aid Overview (2002–2023)",
       subtitle = "Solid lines: Aid & Tuition | Dashed line: % Receiving Aid",
       x = "Year", y = "Amount in Dollars",
       color = "Metric") +
  scale_color_manual(values = c(
    "Average Aid" = "#006400",    
    "Total Tuition" = "#002366",    
    "Percent Given Aid" = "#c9b1c9" 
  )) +
  theme_minimal(base_size = 14) +
  scale_y_continuous(labels = label_dollar(prefix = "$", big.mark = ","))

interactive_plot <- ggplotly(p, tooltip = "text")

interactive_plot




```

Additionally, the university expanded their financial aid program to meet all demonstrated need for first-year domestic students in the fall of 2020. Interestingly, there is limited change in the percent of students given aid and the average aid amount. The average aid number represents "federal, state, local or institutional grant aid awarded to full-time first-time undergraduates," so while full-demonstrated need is an excellent guarantee for students, it appears this was not a financial investment on the university and on average students did not benefit.

# Racial Demographics

In 2009, the race categorizations changed. This affirms that race is a social construct that groups people arbitrarily— NOT based on scientific, genetic, or cultural facts.

The data between 1994 and 2007 group students into the following race categories:

-   White Non-Hispanic

-   Black Non-Hispanic

-   Hispanic

-   Asian or Pacific Islander

-   American Indian or Alaska Native

-   Race/Ethnicity Unknown

There was a transition in categorization in 2008 and 2009, followed by a full implementation of the new race/ethnicity categories in 2010:

-   American Indian or Alaska Native

-   Asian

-   Black or African American

-   Hispanic

-   Native Hawaiian or Other Pacific Islander

-   White

-   Two or More Races

-   Race/Ethnicity Unknown

Interestingly, the 1994 model *leads* with white, while the 2010 model is alphabetical. This is a small, but difficult to ignore detail.

```{r student, message=FALSE, warning=FALSE}

suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(stringr)))
suppressMessages(suppressWarnings(library(tidyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggthemes)))
suppressMessages(suppressWarnings(library(hrbrthemes)))
suppressMessages(suppressWarnings(library(plotly)))
suppressMessages(suppressWarnings(library(scales)))
library(viridis)

data <- read_csv("Stu.Rac.94.csv")

undergrad <- data %>%
  select(Year = `...1`,
         AIAN = `Undergraduate AIAN`,
         Asian = `Undergraduate Asian`,
         Black = `Undergraduate Black`,
         Hispanic = `Undergraduate Hispanic`,
         NHPI = `Undergraduate NHPI`,
                Unknown = `Undergraduate Unknown`,
         White = `Undergraduate White`,
         TwoOrMore = `Undergraduate Two`) %>%
  filter(!is.na(Year)) %>%
  mutate(across(-Year, as.numeric)) %>%
  pivot_longer(-Year, names_to = "Group", values_to = "Count") %>%
  group_by(Year) %>%
  mutate(Percent = Count / sum(Count, na.rm = TRUE) * 100)

UG <- ggplot(undergrad, aes(x = as.integer(Year), y = Percent, fill = Group)) +
  geom_area(alpha = 0.8 , size = 0, colour = "white") +
    scale_fill_viridis_d(option = "viridis") + 
  labs(title = "Undergraduate Racial/Ethnic Composition Over Time",
       x = "Year", y = "Percent", fill = "Group") +
  theme_minimal() +
  theme(legend.position = "right")

ggplotly(UG)


```

```{r grad, message=FALSE, warning=FALSE}

suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(stringr)))
suppressMessages(suppressWarnings(library(tidyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggthemes)))
suppressMessages(suppressWarnings(library(hrbrthemes)))
suppressMessages(suppressWarnings(library(plotly)))
suppressMessages(suppressWarnings(library(scales)))
library(viridis)

data <- read_csv("Stu.Rac.94.csv")

grad <- data %>%
  select(Year = `...1`,
         AIAN = `Graduate AIAN`,
         Asian = `Graduate Asian`,
         Black = `Graduate Black`,
         Hispanic = `Graduate Hispanic`,
         NHPI = `Graduate NHPI`,
                Unknown = `Graduate Unknown`,
         White = `Graduate White`,
         TwoOrMore = `Undergraduate Two`) %>%
  filter(!is.na(Year)) %>%
  mutate(across(-Year, as.numeric)) %>%
  pivot_longer(-Year, names_to = "Group", values_to = "Count") %>%
  group_by(Year) %>%
  mutate(Percent = Count / sum(Count, na.rm = TRUE) * 100)

UG <- ggplot(grad, aes(x = as.integer(Year), y = Percent, fill = Group)) +
  geom_area(alpha = 0.8 , size = 0, colour = "white") +
    scale_fill_viridis_d(option = "viridis") + 
  labs(title = "Graduate Racial/Ethnic Composition Over Time",
       x = "Year", y = "Percent", fill = "Group") +
  theme_minimal() +
  theme(legend.position = "right")

ggplotly(UG)


```

Another observation is that the number of unknown students decreased during this change, implying that the race of these students was never really "Unknown," just unrepresented.

Over time, Boston University's undergraduate and graduate populations have had greater numbers of students from minoritized groups. This more diverse student body enriches our campus and brings new perspectives and knowledge to the classroom and community. However, this dataset only goes to 2023. On October 15th, 2024, BU replaced the "Equal Opportunity/Affirmative Action Policy" in order to comply with the results of the Students for Fair Admissions v. President and Fellows of Harvard College, and Students for Fair Admissions v. University of North Carolina court cases:

*"Boston University recognizes that non-discrimination does not ensure that equal opportunity is a reality. Accordingly, the University will continue to take affirmative action to achieve equal opportunity through recruitment, outreach, and internal reviews of policies and practices."*

As a result of eliminating this policy, the number of black students enrolled in the class of 2028 dropped from 9% to 3%. The United States is at a turning point in higher education, and the data that has yet to be collected will be immensely revealing.

Also, there is a noted difference in the population of the undergraduate vs graduate student body in 2023, where the number of students in graduate programs who identify as Asian, Black, and Hispanic are lower in each category than the proportion of the undergraduate population. This demonstrates the leaky pipeline, where representation decreases as we go up through levels of education.

# International Population

Boston University is the 5th most international school in the United States, with 22% of the total population in 2023 being international students.

-   The New School (34%)

-   New York University (26%)

-   Carnegie Mellon University (23%)

-   University of Rochester (23%)

-   Boston University (22%)

The diverse and global identity of BU has been long standing, and it is an element that the institution prides itself in.

*"BU is already a leader in global engagement, with a deserved reputation as a world-class research university that welcomes gifted students from every corner of the earth."* BU Strategic Plan

```{r studentint, message=FALSE, warning=FALSE}

suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(stringr)))
suppressMessages(suppressWarnings(library(tidyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggthemes)))
suppressMessages(suppressWarnings(library(hrbrthemes)))
suppressMessages(suppressWarnings(library(plotly)))
suppressMessages(suppressWarnings(library(scales)))
library(viridis)

data <- read_csv("Stu.Rac.94.csv")

undergrad <- data %>%
  select(Year = `...1`,
         Total = `Undergraduate Total`,
         International = `Undergraduate International`) %>%
  filter(!is.na(Year)) %>%
  mutate(Domestic = Total - International) %>%
  pivot_longer(cols = c(Domestic, International), names_to = "Category", values_to = "Count") %>%
  group_by(Year) %>%
  mutate(Percent = Count / sum(Count, na.rm = TRUE) * 100)

UG_resident_nonresident <- ggplot(undergrad, aes(x = as.integer(Year), y = Percent, fill = Category)) +
  geom_area(alpha = 0.8, size = 0, colour = "white") +  
  scale_fill_viridis_d(option = "plasma") + 
  labs(title = "Undergrad Domestic vs International Students Over Time",
       x = "Year", y = "Percent", fill = "Category") +
  theme_minimal() +
  theme(legend.position = "right")

ggplotly(UG_resident_nonresident)


```

```{r gradstudentint, message=FALSE, warning=FALSE}

suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(stringr)))
suppressMessages(suppressWarnings(library(tidyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggthemes)))
suppressMessages(suppressWarnings(library(hrbrthemes)))
suppressMessages(suppressWarnings(library(plotly)))
suppressMessages(suppressWarnings(library(scales)))
library(viridis)

data <- read_csv("Stu.Rac.94.csv")

grad <- data %>%
  select(Year = `...1`,
         Total = `Graduate Total`,
         International = `Graduate International`) %>%
  filter(!is.na(Year)) %>%
  mutate(Domestic = Total - International) %>%
  pivot_longer(cols = c(Domestic, International), names_to = "Category", values_to = "Count") %>%
  group_by(Year) %>%
  mutate(Percent = Count / sum(Count, na.rm = TRUE) * 100)

G_resident_nonresident <- ggplot(grad, aes(x = as.integer(Year), y = Percent, fill = Category)) +
  geom_area(alpha = 0.8, size = 0, colour = "white") + 
  scale_fill_viridis_d(option = "plasma") + 
  labs(title = "Graduate Domestic vs International Students Over Time",
       x = "Year", y = "Percent", fill = "Category") +
  theme_minimal() +
  theme(legend.position = "right")

ggplotly(G_resident_nonresident)


```

The number of international students has remained relatively consistent, with marginal increases over time. However, the university will need to adjust as the country becomes more hostile to our international community. To be able to see the trends thus far through data visualization, we will be able to clearly represent the impact that administrative changes from the federal government have on the university in a range of capacities. Yet, the fact that BU is a global, international, and diverse institution cannot be understated.

# Majors

CIP Codes are part of the Classification of Instructional Programs system. There are 61 2 digit codes, and then additional 4 digit codes that provide specificity.

As an example, a BA in Computer Science from BU has the code 11.0701, which 11 marking the Computer and Information Sciences field.

Below, I will analyze the following codes for degree completions each year.

-   09 - Communication and Journalism

-   11 - Computer and Information Sciences

-   13 - Education

-   14 - Engineering

-   16 - Foreign Languages, Literatures, Linguistics

-   23 - English Language and Literature

-   26 - Biological and Biomedical Sciences

-   27 - Mathematics and Statistics

-   38 - Philosophy and Religious Studies

-   40 - Physical Sciences

-   42 - Psychology

-   45 - Social Sciences

-   50 - Visual and Performing Arts

-   52 - Business, Management, Marketing

```{r stumaj, message=FALSE, warning=FALSE}


suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(stringr)))
suppressMessages(suppressWarnings(library(tidyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggthemes)))
suppressMessages(suppressWarnings(library(hrbrthemes)))
suppressMessages(suppressWarnings(library(plotly)))
suppressMessages(suppressWarnings(library(scales)))


data <- read_csv("Stu.CID.01.csv")

bachelor_data <- data %>%
  select(Year, contains("Bachelor"))

bachelor_long <- bachelor_data %>%
  pivot_longer(-Year, names_to = "Group", values_to = "Count") %>%
  separate(Group, into = c("Gender", "Degree", "Field"), sep = "\\. ") %>%
  filter(Degree == "Bachelor")
CIDPLOT <- ggplot(bachelor_long, aes(x = Year, y = Count, color = Gender)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values = c(
  "Women" = "#905acd",
  "Men" = "#20B2AA" 
)) +
  facet_wrap(~ Field, scales = "free_y") +
  labs(title = "Bachelor Degrees by Field and Gender",
       y = "Number of Degrees",
       x = "Year") +
  theme_minimal()+
    theme(
    strip.text = element_text(size = 6)
  )

ggplotly(CIDPLOT)


```

Many degree programs have a significant underrepresentation of men and women. Men are notably underrepresented in the arts, communications/journalism, education, and psychology. Women are underrepresented in computer science and engineering– which are the higher paying fields. There have been pushes in recent years to have more women in STEM fields, but the these graphs do not reveal any notable increase in women where there is not a corresponding increase in men, ie. more people are majoring in that field, not only women. We also see that the number of men is constant in the fields they are underrepresented in.

Computer science, biological and biomedical sciences, and mathematics and statistics are increasing in popularity, while English, Philosophy and Religion, and Education are decreasing in popularity.

In 2018, Boston University and Wheelock College merged together to create a school with the research and doctoral programs from BU's School of Education and the early childhood education and specialized pedagogy of Wheelock in one program. Following the merger, the number of education degrees awarded decreased.

Before the merger, BU offered an array of majors related to education, including a BS in Deaf Studies, BS in Early Childhood Education, BS in Elementary Education, BS in English Education, BS in Mathematics Education, BS in Science Education, BS in Social Studies Education, BS in Special Education, and a BS in Foreign Language Education. In Fall of 2022, BU consolidated all of these majors into the BS in Education & Human Development, which then has tracks based on student interests, and clearly focused in on the Master's and EdD programs in education rather than the Bachelors programs that Wheelock College was known for. Furthermore, the number of Master's degrees awarded in Education remains around five times the number of Bachelor's degrees awarded in Education.

```{r stumas, message=FALSE, warning=FALSE}


suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(stringr)))
suppressMessages(suppressWarnings(library(tidyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggthemes)))
suppressMessages(suppressWarnings(library(hrbrthemes)))
suppressMessages(suppressWarnings(library(plotly)))
suppressMessages(suppressWarnings(library(scales)))


data <- read_csv("Stu.CID.01.csv")

master_data <- data %>%
  select(Year, contains("Master"))

master_long <- master_data %>%
  pivot_longer(-Year, names_to = "Group", values_to = "Count") %>%
  separate(Group, into = c("Gender", "Degree", "Field"), sep = "\\. ") %>%
  filter(Degree == "Master")
MASPLOT <- ggplot(master_long, aes(x = Year, y = Count, color = Gender)) +
  geom_line(size = 1.2) +
scale_color_manual(values = c(
  "Women" = "#905acd",
  "Men" = "#20B2AA" 
)) +
  geom_point(size = 2) +
  facet_wrap(~ Field, scales = "free_y") +
  labs(title = "Master Degrees by Field and Gender",
       y = "Number of Degrees",
       x = "Year") +
  theme_minimal()+
    theme(
    strip.text = element_text(size = 6)
  )

ggplotly(MASPLOT)

```

Many of the same trends are present for Master's Programs, but there are some discrepancies.

One of the notable differences is in the Business, Management, and Marketing CID code. After being renamed from School of Management (SMG) to Questrom School of Business in 2015, the branding and marketing for programs has transitioned, which may explain the visibly exponential increase in Master's students in Business programs. The slogan "Questrom Means Business" is omnipresent in and out of BU, with advertisements on the MBTA and on sponsored Google searches. The Part-Time Flex MBA is growing in popularity, and is something that BU is clearly pushing for more students in, especially since the curriculum is accessible for working professionals!

```{r stutot, message=FALSE, warning=FALSE}


suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(stringr)))
suppressMessages(suppressWarnings(library(tidyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggthemes)))
suppressMessages(suppressWarnings(library(hrbrthemes)))
suppressMessages(suppressWarnings(library(plotly)))
suppressMessages(suppressWarnings(library(scales)))




data <- read_csv("Stu.CID.01.csv")


long_data <- data %>%
  pivot_longer(
    cols = -Year,
    names_to = "Category",
    values_to = "Count"
  ) %>%
  separate(Category, into = c("Gender", "DegreeType", "Field"), sep = "\\. ", extra = "merge") 

long_data_clean <- long_data %>%
  filter(!is.na(Count))

summed_field_year <- long_data_clean %>%
  group_by(Year, Field) %>%
  summarize(Total = sum(Count, na.rm = TRUE)) %>%
  ungroup()

total <- ggplot(summed_field_year, aes(x = Year, y = Total, color = Field)) +
  geom_line(size = 0.5) +
  geom_point(size = 1) +
  labs(title = "Trends in Degrees Awarded by Field",
       x = "Year",
       y = "Total Degrees Awarded") +
  theme_minimal()

ggplotly(total)
```

Many fields have not changed in popularity over time; yet some have increased and decreased with societal pressures.

We see that the stereotypically "high paying" fields are increasing the most. As BU's cost of tuition has increased, the return on investment is a key piece in deciding where to go to college. With a degree that is perceived as less valuable, it is more difficult to justify the high tuition rates. Yet, with a degree in Business, Biology, Engineering, and Computer Science, high salaries are more likely. Computer Science in particular has seen a drastic increase. In 2006, 255 degrees were awarded in Computer Science in 2023, 1093 degrees were awarded. CS is also currently the largest major in the College of Arts & Sciences!

Education is the only field with a noted decrease.

# Conclusion

The student body is diverse. In terms of where they call home, what they study, what they identify as, and what path they are on, BU brings together many different students under the same goal of spreading knowledge.
